---
title: "Gene Set Analysis for time-to-event outcome:comparison of a new approach based on theGeneralized Berk–Jones statistic with existingmethods in presence of intra gene-set correlation"
subtitle: "Supplementary"
author: ""
date: "`r Sys.Date()`"
output:
  bookdown::pdf_document2:
    toc: True
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r}
library(dplyr)
library(ggplot2)

source(file = here::here("functions/fct_report_simu_setting.R"))

dfRembrandt <- readRDS(here::here("data/rembrandt/dfresvarcovar.rds"))
dfBreast <- readRDS(here::here("data/breast_cancer/dfresvarcovar.rds"))
```

# Rembrandt and Breast Cancer Data for Simulation Settings

This section aims to identify realistic simulation settings using the Rembrandt and Breast Cancer datasets.

## Proportion of Significant Genes

As shown in Figure \@ref(fig:propsigngenes), the proportion of significant genes is approximately 45% for the Rembrandt dataset and 20% for the Breast Cancer dataset.

```{r propsigngenes, fig.cap="Proportion of significant genes in each pathway for each data source.", fig.height=4}
dplyr::bind_rows(dfRembrandt$dfCoxVariance |> mutate(data = "Rembrandt"),
                 dfBreast$dfCoxVariance |> mutate(data = "Breast")) |> 
  group_by(pathway, data) |> 
  summarise(nb_genes = n(),
            nb_sign_genes = sum(SIGN),
            prop_sign_genes = nb_sign_genes/nb_genes,
            .groups = "drop") |> 
  ggplot(mapping = aes(y = prop_sign_genes, x = data)) +
  geom_boxplot() +
  scale_y_continuous(breaks = seq(0, 1, by = .2)) +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(x = "Data source",
       y = "Proportion of significant genes")

```

## Gene expression variance

Gene expression variance was studied in each data source.
It was of approximately $0.3$ and $0.09$ in Rembrandt and Breast cancer data respectively.
For simulations, we use an intermediate variance of 0.2 (Figure \@ref(fig:genevariance)).

```{r genevariance, fig.cap="Gene expression variance for each data source.", fig.height=4}
dplyr::bind_rows(dfRembrandt$dfCoxVariance |> mutate(data = "Rembrandt"),
                 dfBreast$dfCoxVariance |> mutate(data = "Breast")) |> 
  select(data, GENE, VARIANCE) |> 
  distinct() |> 
  ggplot(mapping = aes(y = VARIANCE, x = data)) +
  geom_boxplot() +
  scale_y_log10(guide = "axis_logticks") +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(x = "Data source",
       y = "Gene Expression Variance")

```

## Gene expression correlation

Gene expression correlation was evaluated on each pathway of each data source.
Correlation was evaluated overall and stratified by gene significance.
Figure \@ref(fig:overallgenecorrelation) indicates that an overall gene correlation following a non-standard beta distribution(20, 20, min = -1, max = 1) is realistic for both Rembrandt and Breast cancer data. This is denoted as case I.
When stratified by gene significance, figure \@ref(fig:genecorrelationbysign) outlines that a non-standard beta distribution(10, 10, min = -1, max = 1) for significant genes and non-standard beta distribution(25, 25, min = -1, max = 1) for non-significant genes is realistic. This is denoted as case II.

A simpler simulation setting considering 0.2 correlation between significant genes and 0 correlation between non-significant genes was also considered.

```{r overallgenecorrelation, fig.cap="Gene expression correlation for each pathway in each data source.", fig.height=5}
ls_R <- fct_plot_correlation(dfRembrandt, caption = "Rembrandt : Shapes of non-standard beta distribution by type of gene")
ls_B <- fct_plot_correlation(dfBreast, caption = "Breast : Shapes of non-standard beta distribution by type of gene")

plot_R <- ls_R$plot_unique + ggtitle("Rembrandt")
plot_B <- ls_B$plot_unique + ggtitle("Breast")

ggpubr::ggarrange(plot_R, plot_B)
```

```{r genecorrelationbysign, fig.cap="Gene expression correlation for each pathway in each data source stratified by gene significance.", fig.height=6}
plot_R <- ls_R$plot_multiple + ggtitle("Rembrandt")
plot_B <- ls_B$plot_multiple + ggtitle("Breast")

ggpubr::ggarrange(plot_R, plot_B)
```

## Beta from Cox model

```{r}
lsbeta_R <- fct_beta_dist(dfRembrandt, caption = "Rembrandt : Proportion of positive, negative and non significant genes")
lsbeta_B <- fct_beta_dist(dfBreast, caption = "Breast cancer : Proportion of positive, negative and non significant genes")
```

Beta coefficient distributions, both overall and stratified by gene significance, were examined in the two datasets.
Figure \@ref(fig:betacoef) indicates that the overall distribution of effect sizes is well approximated by a normal distribution with mean 0 and standard deviation 0.4; we refer to this setting as Type A.
Figure \@ref(fig:betacoefstrat) further shows that, in the Rembrandt dataset, half of the gene effects follow a normal distribution with mean -0.4 and standard deviation 0.2, while the other half follow a normal distribution with mean 0.4 and standard deviation 0.2; we denote this as Type B.
In contrast, the Breast Cancer dataset is characterized by a wider separation, with half of the effects following a normal distribution with mean -0.8 and standard deviation 0.4, and the other half following a normal distribution with mean 0.8 and standard deviation 0.4; this is referred to as Type C.

```{r betacoef, fig.cap="Beta coefficient distribution for each data source.", fig.height=5}
plot_R <- lsbeta_R$plot_unique_beta + ggtitle("Rembrandt")
plot_B <- lsbeta_B$plot_unique_beta + ggtitle("Breast")

ggpubr::ggarrange(plot_R, plot_B)
```

```{r betacoefstrat, fig.cap="Beta coefficient distribution for each data source stratified by gene significance.", fig.height=6}
plot_R <- lsbeta_R$plot_multiple_beta + ggtitle("Rembrandt")
plot_B <- lsbeta_B$plot_multiple_beta + ggtitle("Breast")

ggpubr::ggarrange(plot_R, plot_B)
```
